Molecular dynamics simulation of synchronization in driven particles

### 2021-01-28 manuscript/manuscript_ajp_2020.tex
updated exercise and table text

\begin{table}[h!]
\centering
\caption{Simulation units and comparable
  experimental values~\cite{Juniper2015}.}
%{\it might be more appropriate to match Juniper parameters more precisely given our focus...}}
\begin{ruledtabular}
\begin{tabular}{c c c } %p{5cm}}
Quantity & Simulation Unit & Experimental range\\
\hline
length &  $a_0 = 1$ & $ a_0 \sim 1.5 \mu m$\\
energy & $E_0 = 1$ & $ < E_0 <$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
%energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
%dimensionless interaction strength & $q$ \\
%effective colloidal charge & $Z^*$ \\
%solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $F_0 = E_0 / a_0$ & $<F_0<$\\
velocity &  $ a_0 / \tau $ &  $v \sim 5 \mu m/s $\\
viscosity coefficient & $\eta = F_0 \tau / a_0 $ & $\eta \sim 10^{-3}$ Pa-s \\
%juniper uses water–ethanol mixture
time &  $ \tau = \eta a_0 / F_0 $ & $ < \tau <$\\
period & $\lambda = 1.825 a_0$ & $\lambda = 3.5 \mu m$ \\
potential amplitude & $U_0 = 0.058 E_0$ & $U_0 = 75 k_B T \sim 2 J$ \\
temperature & $T_0 = 0$ \footnote{a. see Ex.~\ref{ex:brownian} to include temperature.} & $T \sim 300K $ \\

@@

%Volpe AJP Parameters
%We will consider a silica microparticle in water with radius R=1micro m, mass m=11picograms, viscosity eta=0.001 Ns/m^2; $\gamma=6\pi \eta R$,  T=300K, and \tau=0.6\micro s.

@@

In the following section
we include exercises to explore our numerical model,
including the 
to linear drag equation in Ex.~\ref{ex:reynolds}
the equation of motion in Ex.~\ref{ex:n2l},
and 
numerical integration techniques in Ex.~\ref{ex:euler}.
We explore changes in the particle motion
with parameter changes in Ex.~\ref{ex:parameters}.
Methods to visualize the dynamics of the particle motion
are in Ex.~\ref{ex:animation}.
An extended model that includes finite temperature effects
is in Ex.~\ref{ex:brownian}.

@@

\begin{enumerate}
\item {\bf Drag Models and Reynolds Numbers}
  Stokes' law describes the viscous drag on a sphere
  moving at velocity $\vec{v}$ as 
  $\vec{F}^{lin} = -3 \pi \eta D \vec{v}$ 
  where $\eta$ is the dynamic fluid viscosity and 
  $D$ is the particle diameter \cite{Taylor2005}.
  In simulation we
  subsume the constants $3 \pi D$
  such that $\eta \rightarrow 3 \pi D \eta$.
  Often drag forces are
  modeled as a polynomial series
  $\vec{F}^{drag} = -b \vec{v} - c v^2 \hat{v} + ... $.
  Truncating the series to the first term
  is justified by demonstrating the sphere
  has a low Reynolds number  
  $R = D v \rho / \eta$
  where $\rho$ is the fluid density and $v$ the particle's speed.
  When $R$ is small, the quadratic and higher order drag terms
  may be ignored in favor of the linear drag term.
  %
  Using reasonable values for the
  experimental analog of this system, 
  demonstrate the Reynolds number is low
  for this system.

  To first order 
  approximate experimental values for
  a colloid are $D = 1 \mu m$ and $v \sim 1 \mu m /s$.
  %Likewise
  To first order $\eta \sim 10^{-3}$ Pa-s 
  since 
  viscosity of water decreases with temperature
  from $1.3 \times 10^{-3}$ Pa-s at $10^{\circ}$ C to 
  from $0.31 \times 10^{-3}$ Pa-s at $90^{\circ}$ C,
  where 1 Pascal is $1$ N/m$^2$.
  Likewise the liquid density $\rho$ varies with temperature
  but  $\rho \sim 10^3$ kg/cm$^3$ is a reasonable
  first order approximation
  \cite{asce}.

  This model should be familiar to readers
  of a standard classical mechanics text, such as Ref.~\cite{Taylor2005}
  when studying 
  models Millikan's oil drops.

  [{\it could certainly do more with this question to make it more computational.  For instance, plot $R$ as a function of any of the parameters $v$, $D$ and $\rho$, include a temperature dependent $\eta(T)$ in the Brownian system.  Suggest students model something big to imagine their own life at low Reynolds number... fairly off topic...}]

@@

\item{\bf Equation of Motion}
  \label{ex:n2l}

  Newton's second law states that
  the acceleration of a particle $i$
  is proportional to 
  the sum of forces on the particle 
  \begin{equation}
  m_i \vec{a}_i = \sum \vec{F}_i
  \label{eq:n2l}
  \end{equation}
  where the constant of proportionality is the
  inertial mass $m_i$
  The addition of a dissipative force in a dynamical model
  on a colloid is typically modeled
  with a drag force proportional to the particle's velocity
  in the opposite direction of motion 
  $\vec{F}^{drag} = - \eta \vec{v}_i$
  where $\eta$ is the drag coefficient.
  The ratio of $m/\eta$ is %has units of time, % and is 
  known as the momentum relaxation time, %.
  and is small for
  particles with low Reynolds numbers.
  When $m/\eta$ is small,
  particle acceleration can be ignored
  entirely.
  (a) Using Newton's Second Law for
  a small momentum relaxation time, 
  demonstrate that a particle confined to a landscape exerting force
  $F^l(\vec{r}_i)$ subject to a time dependent drive $F^d(t)$
  can be modeled with the equation of motion described in 
  Eq.~\ref{eq:motion}.

  {\it this isn't very challenging, but I've see undergraduate students have a ``wow'' factor when they work this out for the first time.  This is closely related to the next exercise which is also primarily analytic rather than computational}

  \item {\bf Integration Methods}
    \label{ex:euler}
    To calculate the position of the particle we
    integrate the equation of motion using
    the standard definition of velocity
    $\vec{v}_i = d\vec{r}_i/dt$ 
    via the 
    %In our model we use the
    Euler method. % to solve this equation.
    Our equation of motion provides
    a direct calculation for particle velocity.
    %
  When solving ordinary differential equations,
  the Euler method is effective for solving linear equations
  of the form
  $dy/dt = f(t,y(t))$ with initial condition $y(t_0) = y_0$.
  The solution is calculated algorithmically
  by stepping in time through $n$ integer steps
  $t_n = t_0 + n \Delta t$.
  At each subsequent step the new
  value for $y$ is calculated as a map solution using
  discrete times 
  $y_{n+1} = y_n + f(t_n,y_n)$.
  The Euler method can be applied to non-linear
  equations effectively if the time step $\Delta t$
  is kept sufficiently small \cite{Newman}.
  In our simulations we use the timestep $\Delta t = 0.001$,
  but we find for a single particle
  we can just as easily used 
  %examine 
  %for
  timesteps $\Delta t = 0.01$ and $\Delta t = 0.1$
  over the same maximum time
  (i.e. decrease the overall simulation time and sampling rate)
  and reproduce the motions shown in Fig.~\ref{fig:0}(b).
  This is due to the linear nature of the solution,
  and the lack of interaction with neighboring particles.
  Apply the Euler method to our equation
  of motion to solve for the analytic expression
  of position of a particle
  $y_n$ at the $nth$ timestep (i.e. Eq.~\ref{eq:euler}).
  %How to the motions of the particle compare when the timestep
  %becomes large?

@@

  Often molecular dynamics algorithms 
  are solved with higher order methods
  such as the Verlet method or Runge-Kutta methods.
  These methods include  higher order terms
  that provide accurate solutions for
  second order differential equations,
  i.e. $dy/dt = v$ and $dv/dt = f_2(y,t)$,
  where $f_2(y,t)$ is a second derivative function.
  In Newton's second law this would be proportional
  to the net force on a single particle.
  We do not integrate $f_2(y,t)$ since 
  $a = dv/dt = 0$,
  and thus rearrange $f_2(y,t)$ to solve directly for $v$,
  as described in Ex.~\ref{ex:n2l}.
  The Taylor expansion of the position function $y(t)$
  \begin{equation}
    y(t_n \pm \Delta t) = y(t_n) \pm \frac{dy}{dt} \Delta t + \frac{1}{2} \frac{d^2y}{dt^2} (\Delta t)^2 \pm O(\Delta t)^3 ... %\frac{1}{6} \frac{d^3y}{dt^3} (\Delta t)^3 + ...
  \end{equation}
  is exact at first order
  for the $n$-th time step.

  Demonstrate how the Verlet method
  simplifies to the Euler method when
  the particle acceleration $\vec{a}=0$.

@@

\item {\bf Brownian motion}
  \label{ex:brownian}
  is a temperature and viscosity dependent
  phenomena. \cite{}
  Consider the consequences of a non-zero temperature on the simulation, in which particles will collide with invisible particles making up the suspending fluid.  These collisions are more likely at higher temperature due to the increased kinetic energy of the fluid particles, as described by the Maxwell-Boltzmann distribution \cite{}.  In molecular dynamics simulations temperature effects can be modeled by including randomized forces $f^T$ to emulate the motion of particles undergoing Brownian motion.  A distribution of randomizes forces contributes equally in all directions such that the force averaged over a finite time interval is zero $\langle f^T(t) \rangle = 0$.  Random distributions of contain no correlations over the independent variable (time in this case) $\langle f^T(t) f^T(\tau)\rangle = 2 \eta k_B T \delta(t-\tau)$ where the energy $k_B T$ derived from the Boltzmann constant $k_B$ and temperature $T$ has energy units $E_0$.  \cite{Allen2017}

  With no applied driving force, measure the minimum temperature required for a single particle to hop between minima in the potential landscape.

  Measure the rate of diffusion.

\item {\bf Exploring Parameters of the Model}
  \label{ex:parameters}

@@

\item {\bf Visualizing motion + phase plots}
  \label{ex:animation}

@@

\end{enumerate}

@@

>>>>

\begin{table}[h!]
\centering
\caption{Simulation parameters and units and comparable
  experimental values~\cite{Juniper2015}.
  Blanks %in the experimental parameters
  indicate
  units, rather than parameters that
  are not easily represented by a single value.}
%{\it might be more appropriate to match Juniper parameters more precisely given our focus...}}
\begin{ruledtabular}
\begin{tabular}{c c c } %p{5cm}}
Quantity & Simulation Unit & Experimental value\\
\hline
length &  $a_0 = 1$ & $ a_0 \sim 1.5 \mu m$\\
energy & $E_0 = 1$ & \\ %$ < E_0 <$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
%energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
%dimensionless interaction strength & $q$ \\
%effective colloidal charge & $Z^*$ \\
%solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $F_0 = E_0 / a_0$ & \\ %$<F_0<$\\
time &  $ \tau = \eta a_0 / F_0 $ \\ %& $ < \tau <$\\
velocity &  $ a_0 / \tau $ &  $v \sim 5 \mu m/s $\\
driving period & $T = 100 \tau$ & $T \sim 1.3$ s \\
viscosity coefficient & $\eta = F_0 \tau / a_0 $ & $\eta \sim 10^{-3}$ Pa-s \\
%juniper uses water–ethanol mixture
substrate period & $\lambda = 1.825 a_0$ & $\lambda = 3.5 \mu m$ \\
substrate amplitude & $U_0 = 0.058 E_0$ & $U_0 = 75 k_B T \sim 2 J$ \\
temperature & $T_0 = 0$ \footnote{see Ex.~\ref{ex:brownian} to include temperature in the limit where $U_0 \sim k_B T$.} & $T \sim 300K $ \\

@@

%Volpe AJP Parameters
%We will consider a silica microparticle in water with radius R=1micro m, mass m=11picograms, viscosity eta=0.001 Ns/m^2; $\gamma=6\pi \eta R$,  T=300K, and \tau=0.6\micro s.
%thermofisher off the shelf paramagnetic beads
%https://www.thermofisher.com/us/en/home/references/protocols/proteins-expression-isolation-and-analysis/protein-isolation-protocol/m-270-carboxylic-acid.html
%~2\times10^9 beads per 30 mg -> 1.5 \times 10^-11 g, 15 ng

@@

In the following section
we include exercises to explore our numerical model,
including the 
to linear drag equation in Ex.~\ref{ex:reynolds}, 
the equation of motion in Ex.~\ref{ex:n2l},
and 
numerical integration techniques in Ex.~\ref{ex:euler}.
An extended model that includes finite temperature effects
is in Ex.~\ref{ex:brownian}.
We explore changes in the particle motion
with parameter changes in Ex.~\ref{ex:parameters}.
Methods to visualize the dynamics of the particle motion
are in Ex.~\ref{ex:animation}.

@@

%\begin{enumerate}
%\item
\subsection{Drag Models and Reynolds Numbers.}
  Stokes' law describes the viscous drag on a sphere
  moving at velocity $\vec{v}$ as 
  $\vec{F}^{lin} = -3 \pi \eta D \vec{v}$ 
  where $\eta$ is the dynamic fluid viscosity and 
  $D$ is the particle diameter \cite{Taylor2005}.
  In simulation we
  subsume the constants $3 \pi D$
  such that $3 \pi D \eta \rightarrow \eta $.
  %
  Often drag forces are
  modeled as a polynomial series ~\cite{Taylor2005}
  \begin{equation}
    \vec{F}^{drag} = -b \vec{v} - c v^2 \hat{v} + \ldots  .
  \end{equation}
  %This model should be familiar to readers
  %of a standard classical mechanics text, such as Ref.
  %when studying 
  %models Millikan's oil drops.
  Truncating the series to the first term
  is justified by demonstrating the sphere
  has a low Reynolds number  
  $R = D v \rho / \eta$
  where $\rho$ is the fluid density and $v$ the particle's speed.
  When $R$ is small, the quadratic and higher order terms
  may be ignored in favor of the linear drag term.

  Using reasonable values for the
  experimental analog of this system, 
  demonstrate the Reynolds number is low.
  In addition to the values listed in Table ~\ref{tab:1}, 
 % Approximate experimental values for
 % a colloid size and speed
 % are $D = 1 \mu m$ and $v \sim 1 \mu m /s$.
 % %Likewise
 % To first order the viscosity is $\eta \sim 10^{-3}$ Pa-s 
 % since 
 % viscosity of water decreases with temperature
 % from $1.3 \times 10^{-3}$ Pa-s at $10^{\circ}$ C to 
 % from $0.31 \times 10^{-3}$ Pa-s at $90^{\circ}$ C,
 % where 1 Pascal is $1$ N/m$^2$.
  %Likewise
  the liquid density $\rho$ varies with temperature
  with $\rho \sim 10^3$ kg/cm$^3$ a reasonable
  first order approximation
  \cite{asce}.

%  [{\it could certainly do more with this question to make it more computational.  For instance, plot $R$ as a function of any of the parameters $v$, $D$ and $\rho$, include a temperature dependent $\eta(T)$ in the Brownian system.  Suggest students model something big to imagine their own life at low Reynolds number... fairly off topic...}]

@@

%\item
\subsection{Equation of Motion}
  \label{ex:n2l}

  Newton's second law states that
  the acceleration of a particle $i$
  is proportional to 
  the sum of forces on the particle 
  \begin{equation}
  m_i \vec{a}_i = \sum \vec{F}_i
  \label{eq:n2l}
  \end{equation}
  where the constant of proportionality is the
  inertial mass $m_i$.  
  The addition of a dissipative force to a dynamical equation 
  of colloid motion 
  is typically modeled
  with a drag force proportional to the particle's velocity
  in the opposite direction of motion 
  $\vec{F}^{drag} = - \gamma \vec{v}_i$
  where $\gamma = 3 \pi \eta D$ is the drag coefficient
  described in Ex.~\ref{ex:reynolds}.
  The ratio of $m/\gamma$ is %has units of time, % and is 
  known as the momentum relaxation time, %.
  and is small for
  particles with low Reynolds numbers.
  The mass of the
  typical experimental colloid particle is $15$ picograms,
  leading to a momentum relaxation time
  % \tau = 15 nanograms/(3 pi D \eta)
  %$15*10^-12 / (3pi * 3*10^-6 meters * 0.001 Pa-s * 1000 g/kg)$  %1 Pa-s = 1 N s /m^2 = 1 kg /(m s)
  on the order of microseconds.
  %of $1.6 ~\mu$s.
  Confirm for the values listed in Table ~\ref{tab:1},
  the momentum relaxation time is 
  $m/\gamma = 0.5 ~\mu$s.

  When $m/\gamma$ is small,
  particle acceleration can be ignored
  entirely.
  (a) Using Newton's Second Law for
  a small momentum relaxation time, 
  demonstrate that a particle confined to a landscape exerting force
  $F^l(\vec{r}_i)$ subject to a time dependent drive $F^d(t)$
  can be modeled with the equation of motion described in 
  Eq.~\ref{eq:motion}.

%  {\it this isn't very challenging, but I've see undergraduate students have a ``wow'' factor when they work this out for the first time.  This is closely related to the next exercise which is also primarily analytic rather than computational}

%\item
  \subsection{Integration Methods}
    \label{ex:euler}
    To calculate the position of the particle we
    integrate the equation of motion using
    the standard definition of velocity
    $\vec{v}_i = d\vec{r}_i/dt$ 
    via the 
    %In our model we use the
    Euler method. % to solve this equation.
    Our equation of motion provides
    a direct calculation for particle velocity,
    as demonstrated in Ex.~\ref{ex:n2l}.

  When solving ordinary differential equations,
  the Euler method is effective for solving linear equations
  of the form
  $dy/dt = f(t,y(t))$ with initial condition $y(t_0) = y_0$.
  The solution is calculated algorithmically
  by stepping in time through $n$ integer steps
  $t_n = t_0 + n \Delta t$.
  At each subsequent step the new
  value for $y$ is calculated as a map solution using
  discrete times 
  $y_{n+1} = y_n + f(t_n,y_n)$.
  The Euler method can be applied to non-linear
  equations effectively if the time step $\Delta t$
  is kept sufficiently small \cite{Newman}.
  In our simulations we use the timestep $\Delta t = 0.001$,
  but we find for a single particle
  we can just as easily used 
  %examine 
  %for
  timesteps $\Delta t = 0.01$ and $\Delta t = 0.1$
  over the same maximum time
  (i.e. decrease the overall simulation time and sampling rate)
  and reproduce the motions shown in Fig.~\ref{fig:0}(b).
  This is due to the linear nature of the solution,
  and the lack of interaction with neighboring particles.
  Apply the Euler method to our equation
  of motion to solve for the analytic expression
  of position of a particle
  $y_n$ at the $nth$ timestep (i.e. Eq.~\ref{eq:euler}).
  %How to the motions of the particle compare when the timestep
  %becomes large?

@@

  Often molecular dynamics algorithms 
  are solved with higher order methods
  such as the Verlet method or Runge-Kutta methods.
  These methods include  higher order terms
  that provide accurate solutions for
  second order differential equations,
  i.e. $dy/dt = v$ and $dv/dt = f_2(y,t)$.
  %where $f_2(y,t)$ is a second derivative function.
  In Newton's second law $f_2$ would be proportional
  to the net force on a single particle.

  In our model 
  we do not integrate $f_2(y,t)$ since 
  $a = dv/dt = 0$,
  and thus rearrange $f_2(y,t)$ to solve directly for $v$,
  as described in Ex.~\ref{ex:n2l}.
  The Taylor expansion of the position function $y(t)$
  \begin{equation}
    y(t_n \pm \Delta t) = y(t_n) \pm \frac{dy}{dt} \Delta t + \frac{1}{2} \frac{d^2y}{dt^2} (\Delta t)^2 + \ldots
    %\pm O(\Delta t)^3 ... %\frac{1}{6} \frac{d^3y}{dt^3} (\Delta t)^3 + ...
  \end{equation}
  is exact at first order
  for the $n$-th time step.

  Demonstrate how the Verlet method
  simplifies to the Euler method when
  the particle acceleration $a=0$.

@@

%\item
  \subsection{Brownian motion}
  \label{ex:brownian}
  Brownian motion 
  is a temperature and viscosity dependent
  phenomena.
  Particles change direction %randomly
  due to collisions with fluid particles.
  The frequency of collisions depends on the properties of
  the suspending fluid.
  These collisions are more likely at higher temperature
  due to the increased kinetic energy of the fluid particles,
  as described by the Maxwell-Boltzmann distribution \cite{Einstein1905}.

  %Consider a non-zero temperature in the simulation.
  %in which particles will collide with invisible particles
  %making up the suspending fluid.
  In molecular dynamics simulations temperature effects
  can be modeled by including randomized forces $f^T$
  to emulate the motion of particles undergoing Brownian motion.
  A distribution of random forces contributes equally in all
  directions such that the force averaged over a finite time interval
  is zero $\langle f^T(t) \rangle = 0$.
  Random distributions are not correlated 
  over the independent variable,
  where we take our independent variable to be time 
  $\langle f^T(t) f^T(\tau)\rangle = 2 \eta k_B T \delta(t-\tau)$
  for a system with 
  energy $k_B T$
  derived from the Boltzmann constant $k_B$
  and temperature $T$ \cite{Allen2017}.
  A particle
  with energy $k_b T_{min} > U_0$ may 
  hop over landscape
  barriers.

  % (a) 
  In simulation we define our temperature parameter
  $k_b T/E_0 \rightarrow T$
  since we set the constants to unity.
  This allows us to choose
e  a temperature that is easily compared to
  the other system parameters that scale the forces.
  For instance, 
  with
  no applied driving force,
  the minimum temperature required for a single particle
  to hop between minima in the potential landscape is
  $T_{min} > A_p$.
  Assuming a particle is confined to
  motion along the $y-$direction, 
  measure the position vs. time
  of a non-driven colloid for $T/A_p = 0.5, 1.0$, and $2.0$
  when Brownian motion is included
  in the model.
  Confirm that a particle will hop a barrier
  for sufficiently high $T/A_p$ ratios.

  A quantity of interest
  for a particle undergoing Brownian motion
  is the rate of displacement
  of the particle from its initial position 
  This is measured as mean square displacement
  as a function of time
  \begin{equation}
    \langle \Delta \vec{r}(t) ^2 \rangle = \langle (\vec{r}(t + t_0) - \vec{r}(t_0))^2\rangle 
  \end{equation}
  where $t_0$ is the initial time at
  initial position $\vec{r}(t_0)$.
  With no substrate, we would expect
  the root mean square displacement (RMSD)
  to be proportional to the elapsed time 
  \begin{equation}
    \sqrt {\langle \Delta \vec{r}(t) ^2 \rangle} \propto t.
  \end{equation}
  Confirm this result for $A_p = 0$.
  Measure the RMSD as a function of time
  of a non-driven colloid for $T/A_p = 0.5, 1.0$, and $2.0$.
  Compare your result to the
  substrate free RMSD.

  %(b)
  How does Brownian motion affect
  the formation of Shapiro steps?
  With driving force reproduce the paper figures...

%\item
  \subsection{Exploring Parameters of the Model}
  \label{ex:parameters}

  La la la...

@@

%\item
  \subsection{Visualizing motion + phase plots}
  \label{ex:animation}

@@

%\end{enumerate}

@@

  \bibitem{Einstein1905} A Einstein, Investigations on the Theory of the Brownian Movement,  Dover Publications (1956).
